3.7: Regression with Interaction Effects - R Practice

### 2019-11-19 content/assignment/paper-project.Rmd
Update paper deadlines and data sources

@@

- National and State Accounts Data: [Bureau of Economic Analysis](http://www.bea.doc.gov/)
- Labor Market and Price Data: [Bureau of Labor Statistics](http://www.bls.gov/)
- Macroeconomic Data: [Federal Reserve Economic Data (FRED)](https://fred.stlouisfed.org/), [World Development Indicators (World Bank)](https://datacatalog.worldbank.org/dataset/world-development-indicators), [Penn World Table](https://www.rug.nl/ggdc/productivity/pwt/)
- International Data: [NationMaster.com](http://www.nationmaster.com/), [Doing Business](http://rru.worldbank.org/DoingBusiness/), [CIESIN](ttp://www.ciesin.org/data.html)
- Census Data: [U.S. Census Bureau](http://www.census.gov/)
- Sports Data: [Spotrac](http://spotrac.com), [Rodney Fort's Sports Data](https://sites.google.com/site/rodswebpages/codes)
- Data Clearing House: [Stat USA](http://www.stat-usa.gov/), [Fedstats](http://www.fedstats.gov/), [Statistical Abstract of the United States](http://www.census.gov/prod/www/statistical-abstract-us.html), [Resources for Economists](http://rfe.org) 
- Political and Social Data: [ICPSR](http://www.icpsr.umich.edu/), [Federal Election Commission](http://fec.gov), [Poole and Rosenthal Roll Call Data (Voting ideology)](http://voteview.uh.edu/default_nomdata.htm), [Archigos Data on Political Leaders](https://www.rochester.edu/college/faculty/hgoemans/data.htm), [Library of Congress: Thomas (Legislation)](http://thomas.loc.gov/), [Iowa Electronic Markets (Prediction Markets)](http://www.biz.uiowa.edu/iem/)
- War and Violence Data: [Correlates of War](http://www.correlatesofwar.org/data-sets)
- State Level Data: [Correlates of State Policy](http://ippsr.msu.edu/public-policy/correlates-state-policy)
- Health Data: [Centers for Disease Control](http://www.cdc.gov/scientific.htm), [CDC Wonder System](http://wonder.cdc.gov/DataSets.shtml)
- Crime Data: [Bureau of Justice Statistics](http://www.ojp.usdoj.gov/bjs/)
- Education Data: [National Center for Education Statistics](http://www.nces.ed.gov/)
- Environmental Data: [EPA](http://www.epa.gov/ceisweb1/ceishome/)
- Religion Data: [American Religion Data Archiva (ARDA)](http://www.arda.tm/)
- Financial Data: [Financial Data Finder](http://www.cob.ohio-state.edu/cgi-bin/DB_Search/db_search.cgi?setup_file=finance.setup.cgi){Financial Data Finder}
- Philanthropy Data: [The Urban Institute](http://nccs.urban.org/)

@@

| Assignment | Points | Due Date | Description |
|------------|--------|----------|-------------|
| Abstract | 5 | October 29 | Short summary of your ideas |
| Data Description | 10 | November 14 | Description of data sources, and some summary statistics |
| Literature Review | 10 | December 5 | 1-3 paragraphs on 2-3 scholarly sources |
| Presentation | 5 | December 10/12 | Short presentation of your project so far |
| Final Paper Due | 70 | December 17 | Email to me paper, data, code |

>>>>

**Good R Packages for Getting Data in R Format**^[Some of these come from [Nick Huntington-Klein's excellent list](http://www.nickchk.com/econometrics.html).]

Below are packages written by and for R users that link up with the API of key data sets for easy use in R. Each link goes to the documentation and description of each package.

Don't forget to install^[`install.packages("name_of_package")`] first and then load it with `library()`.

- [`wbstats`](https://cran.r-project.org/web/packages/wbstats/wbstats.pdf) provides access to all the data available on the **World Bank** API, which is basically everything on their website. The World Bank keeps track of *many* country-level indicators over time.
- [`tidycensus`](https://cran.r-project.org/web/packages/tidycensus/tidycensus.pdf) gives you access to data from the **US Census** and the **American Community Survey.** These are the largest high-quality data sets you'll find of cross-sectional data on individual people in the US. You'll need to get a (free) API key from the website (or ask me for mine).
- [`fredr`](https://cran.r-project.org/web/packages/fredr/vignettes/fredr.html) gets data from the **Federal Reserve's Economic Database (FRED)**. You'll need to get a (free) API key from the website (or ask me for mine).
- [`tidyquant`](https://cran.r-project.org/web/packages/tidyquant/index.html) gets data from a number of financial sources (including `fredr).
- [`icpsrdata`](https://cran.r-project.org/web/packages/icpsrdata/icpsrdata.pdf) downloads data from the **Inter-university Consortium for Political and Social Research** (you'll need an account and a keycode). ICPSR is a database of datasets from published social science papers for the purposes of reproducibility.
- [`NHANES`](https://cran.r-project.org/web/packages/NHANES/NHANES.pdf) uses data from the **US National Health and Nutrition Examination Survey.**
- [`ipumsr`](https://cran.r-project.org/web/packages/ipumsr/vignettes/ipums.html) has **census data** from *all around the world*, in addition to the US census, **American Community Survey**, and **Current Population Survey**. If you're doing international micro work, look at IPUMS. It's also the easiest way to get the **Current Population Survey (CPS)**, which is very popular for labor economics. Unfortunately ipumsr won't get the data from within R; you'll have to make your own data extract on the IPUMS website and download it. But ipumsr will read that file into R and preserve things like names and labels.
- [`education-data-package-r`](https://github.com/UrbanInstitute/education-data-package-r)^[Note you will need to install `devtools` package first, and then install the package directly from Github with the command `devtools::install_github('UrbanInstitute/education-data-package-r')`] is the Urban Institute's data data on educational institutions in the US, including colleges (in IPEDS) and K-12 schools (in CCD). This package also has data on county-level poverty rates from SAIPE.
- [`psidR`](https://cran.r-project.org/web/packages/psidR/psidR.pdf) is the **Panel Study of Income Dynamics**. This study doesn't just follow people over their lifetimes, it follows their children too, generationally! A great source for studying how things follow families through generations.
- [`atus`](https://cran.r-project.org/web/packages/atus/atus.pdf) is th e **American Time Use Survey**, which is a large cross-sectional data set with information on how people spend their time.
- [`Rilostat`](https://ilostat.github.io/Rilostat/index.html) uses data from the **International Labor Organization.** This contains lots of different statistics on labor, like employment, wage gaps, etc., generally aggregated to the national level and changing over time.
- [`democracyData`](https://github.com/xmarquez/democracyData)^[Note you will need to install `devtools` package first, and then install the package directly from Github with the command `devtools::install_github('xmarquez/democracyData')`] is a great "package for accessing and manipulating existing measures of democracy."
- [`politicaldata`](https://cran.rstudio.com/web/packages/politicaldata/index.html) provides useful functions for obtaining commonly-used data in political analysis and political science, including from sources such as the Comparative Agendas Project (which provides data on politics and policy from 20+ countries), the MIT Election and Data Science Lab, and FiveThirtyEight.

@@

- [Quality of Government Data](https://qog.pol.gu.se/data/dataextras/datasources) has an extremely wide range of data sources pertaining to measures of institutions. The data itself can be found [here](https://qog.pol.gu.se/data/datadownloads/qogstandarddata). 
- National and State Accounts Data: [Bureau of Economic Analysis](http://www.bea.doc.gov/)
- Labor Market and Price Data: [Bureau of Labor Statistics](http://www.bls.gov/)
- Macroeconomic Data: [Federal Reserve Economic Data (FRED)](https://fred.stlouisfed.org/), [World Development Indicators (World Bank)](https://datacatalog.worldbank.org/dataset/world-development-indicators), [Penn World Table](https://www.rug.nl/ggdc/productivity/pwt/)
- International Data: [NationMaster.com](http://www.nationmaster.com/), [Doing Business](http://rru.worldbank.org/DoingBusiness/), [CIESIN](ttp://www.ciesin.org/data.html)
- Census Data: [U.S. Census Bureau](http://www.census.gov/)
- Sports Data: [Spotrac](http://spotrac.com), [Rodney Fort's Sports Data](https://sites.google.com/site/rodswebpages/codes)
- Data Clearing House: [Stat USA](http://www.stat-usa.gov/), [Fedstats](http://www.fedstats.gov/), [Statistical Abstract of the United States](http://www.census.gov/prod/www/statistical-abstract-us.html), [Resources for Economists](http://rfe.org) 
- Political and Social Data: [ICPSR](http://www.icpsr.umich.edu/), [Federal Election Commission](http://fec.gov), [Poole and Rosenthal Roll Call Data (Voting ideology)](http://voteview.uh.edu/default_nomdata.htm), [Archigos Data on Political Leaders](https://www.rochester.edu/college/faculty/hgoemans/data.htm), [Library of Congress: Thomas (Legislation)](http://thomas.loc.gov/), [Iowa Electronic Markets (Prediction Markets)](http://www.biz.uiowa.edu/iem/)
- War and Violence Data: [Correlates of War](http://www.correlatesofwar.org/data-sets)
- State Level Data: [Correlates of State Policy](http://ippsr.msu.edu/public-policy/correlates-state-policy)
- Health Data: [Centers for Disease Control](http://www.cdc.gov/scientific.htm), [CDC Wonder System](http://wonder.cdc.gov/DataSets.shtml)
- Crime Data: [Bureau of Justice Statistics](http://www.ojp.usdoj.gov/bjs/)
- Education Data: [National Center for Education Statistics](http://www.nces.ed.gov/)
- Environmental Data: [EPA](http://www.epa.gov/ceisweb1/ceishome/)
- Religion Data: [American Religion Data Archiva (ARDA)](http://www.arda.tm/)
- Financial Data: [Financial Data Finder](http://www.cob.ohio-state.edu/cgi-bin/DB_Search/db_search.cgi?setup_file=finance.setup.cgi){Financial Data Finder}
- Philanthropy Data: [The Urban Institute](http://nccs.urban.org/)

@@

| Assignment | Points | Due Date | Description |
|------------|--------|----------|-------------|
| Abstract | 5 | October 29 | Short summary of your ideas |
| Data Description | 10 | November 24 | Description of data sources, and some summary statistics |
| Literature Review | 10 | December 6 | 1-3 paragraphs on 2-3 scholarly sources |
| Presentation | 5 | December 5/10 | Short presentation of your project so far |
| Final Paper Due | 70 | December 17 | Email to me paper, data, code |

### 2019-11-19 content/class/18-class.Rmd
Update paper deadlines and data sources

[Problm Set 4](/assignment/04-problem-set) (on classes 3.1-3.5) is due by Tuesday November 19.

>>>>

[Problm Set 4](/assignment/04-problem-set) (on classes 3.1-3.5) is due by Thursday November 21.

# Appendix: T-Test for Difference in Group Means

Often we want to compare the means between two groups, and see if the difference is statistically significant. As an example, **is there a statistically significant difference in average hourly earnings between men and women**? Let:

- $\mu_W$: mean hourly earnings for female college graduates
- $\mu_M$: mean hourly earnings for male college graduates

We want to run a hypothesis test for the difference $(d)$ in these two population means: 
$$\mu_M-\mu_W=d_0$$

Our null hypothesis is that there is *no* statistically significant difference. Let's also have a two-sided alternative hypothesis, simply that there *is* a difference (positive or negative).

- $H_0: d=0$
- $H_1: d \neq 0$

Note a logical one-sided alternative would be $H_2: d > 0$, i.e. men earn more than women

### The Sampling Distribution of $d$

The *true* population means $\mu_M, \mu_W$ are unknown, we must estimate them from *samples* of men and women. Let:
- $\bar{Y}_M$ the average earnings of a sample of $n_M$ men  
- $\bar{Y}_W$ the average earnings of a sample of $n_W$ women

We then estimate $(\mu_M-\mu_W)$ with the sample $(\bar{Y}_M-\bar{Y}_W)$.

We would then run a **t-test** and calculate the **test-statistic** for the difference in means. The formula for the test statistic is:

$$t = \frac{(\bar{Y_M}-\bar{Y_W})-d_0}{\sqrt{\frac{s_M^2}{n_M}+\frac{s_W^2}{n_W}}}$$

We then compare $t$ against the critical value $t^*$, or calculate the $p$-value $P(T>t)$ as usual to determine if we have sufficient evidence to reject $H_0$

```{r}
library(tidyverse)
library(wooldridge)
# Our data comes from wage1 in the wooldridge package

wages<-wooldridge::wage1

# look at average wage for men

wages %>%
  filter(female==0) %>%
  summarize(average = mean(wage),
            sd = sd(wage))

# look at average wage for women

wages %>%
  filter(female==1) %>%
  summarize(average = mean(wage),
            sd = sd(wage))

```

So our data is telling us that male and female average hourly earnings are distributed as such:

$$\begin{align*}
\bar{Y}_M &\sim N(7.10,4.16)\\
\bar{Y}_W &\sim N(4.59,2.53)\\
\end{align*}$$

We can plot this to see visually. There is a lot of overlap in the two distributions, but the male average is higher than the female average, and there is also a lot more variation in males than females, noticeably the male distribution skews further to the right.

```{r}
wages$female<-as.factor(wages$female)

library("ggplot2")
ggplot(data=wages,aes(x=wage,fill=female))+
  geom_density(alpha=0.5)+
  scale_x_continuous(seq(0,25,5),name="Wage",labels=scales::dollar)+
  theme_light()
```

Knowing the distributions of male and female average hourly earnings, we can estimate the **sampling distribution of the difference in group eans** between men and women as:

The mean: 
$$\begin{align*}
\bar{d}&=\bar{Y}_M-\bar{Y}_W\\
\bar{d}&=7.10-4.59\\
\bar{d}&=2.51\\
\end{align*}$$

The standard error of the mean:
$$\begin{align*}
SE(\bar{d})&=\sqrt{\frac{s_M^2}{n_M}+\frac{s_W^2}{n_W}}\\
&=\sqrt{\frac{4.16^2}{274}+\frac{2.33^2}{252}}\\
& \approx 0.29\\
\end{align*}$$

So the sampling distribution of the difference in group means is distributed:
$$\bar{d} \sim N(2.51,0.29)$$

```{r}
ggplot(data.frame(x=0:6),aes(x=x))+
  stat_function(fun=dnorm, args=list(mean=2.51, sd=0.29), color="purple")+
  ylab("Density")+
  scale_x_continuous(seq(0,6,1),name="Wage Difference",labels=scales::dollar)+
  theme_light()
```

Now we the **$t$-test** like any other:

$$\begin{align*}
t&=\frac{\text{estimate}-\text{null hypothesis}}{\text{standard error of the estimate}}\\
&=\frac{d-0}{SE(d)}\\
&=\frac{2.51-0}{0.29}\\
&=8.66\\
\end{align*}$$

This is statistically significant. The $p$-value, $P(t>8.66)=$ is 0.000000000000000000410, or basically, 0.

```{r}
pt(8.66,456.33, lower.tail=FALSE)
```

### The $t$-test in `R`

```{r}
t.test(wage~female, data=wages, var.equal=FALSE)
```

```{r}
reg<-lm(wage~female, data=wages)
summary(reg)
```

### 2019-11-19 content/class/19-class.Rmd
Update paper deadlines and data sources

>>>>

Another way to visualize what we've been covering:

```{r, fig.retina=3}
library(mosaic)

control=function(x){2+0.5*x}
treat=function(x){4+x}

ggplot(data.frame(x=c(0,10)), aes(x=x))+
  stat_function(fun=control, geom="line", size=2, color = "red")+
    geom_label(aes(x=5,y=control(5)), color = "red", label=expression(paste("slope=",hat(beta[1]))), size = 5)+
  stat_function(fun=treat, geom="line", size=2, color = "blue")+
    geom_label(aes(x=5,y=treat(5)), color = "blue", label=expression(paste("slope=",hat(beta[1])+hat(beta[3]))), size = 5)+
    scale_x_continuous(breaks=NULL,
                     limits=c(0,10),
                     expand=expand_scale(mult=c(0,0.1)))+
  scale_y_continuous(breaks=c(2,4),
                     labels=c(expression(hat(beta[0])),expression(hat(beta[0])+hat(beta[2]))),
                     limits=c(0,10),
                     expand=expand_scale(mult=c(0,0.1)))+
  labs(x = "",
       y = "")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=20)

```

$$\widehat{\text{wage}}_i=\hat{\beta_0}+\hat{\beta_1} \, \text{experience}_i$$

$$\widehat{\text{wage}}_i=\hat{\beta_0}+\hat{\beta_1} \, \text{experience}_i+\hat{\beta_2} \, \text{gender}_i$$

$$\widehat{\text{wage}}_i=\hat{\beta_0}+\hat{\beta_1} \, \text{experience}_i+\hat{\beta_2} \, \text{gender}_i+ \hat{\beta_3}(\text{experience}_i \times \text{gender}_i)$$

```{r}
library(tidyverse)
library(wooldridge)
# Our data comes from wage1 in the wooldridge package

wages<-wooldridge::wage1
```

```{r, fig.retina=3}
ggplot(data = wages)+
  aes(x = exper,
      y = wage)+
  geom_point()+
  geom_smooth(method="lm")+
  scale_y_continuous(labels=scales::dollar)+
  labs(x = "Experience (years)",
       y = "Wage")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
```

```{r}
library(broom)
gender_reg<-lm(wage~exper+female, data = wages) %>%
  tidy()

gender_reg

slope<-gender_reg %>%
  filter(term=="exper") %>%
  pull(estimate)
slope

male_intercept<-gender_reg %>%
  filter(term=="(Intercept)")%>%
  pull(estimate)

female_intercept<-gender_reg %>%
  filter(term=="female") %>%
  pull(estimate)

female_intercept<-female_intercept+male_intercept
```

```{r, fig.retina=3}
colors = c("1" = "pink", "0" = "blue")

ggplot(data = wages)+
  aes(x = exper,
      y = wage,
      color = as.factor(female))+
  geom_point()+
  geom_abline(slope=slope, intercept = male_intercept, size=2, color = "blue")+
  geom_abline(slope=slope, intercept = female_intercept, size=2, color = "pink")+
  geom_label(x=2.5,y=6.63, color="blue", label="6.63")+
  geom_label(x=2.5,y=4.15, color="pink", label="4.15")+
  scale_y_continuous(labels=scales::dollar)+
  scale_color_manual("Gender", labels=c("Men","Women"), values = colors)+
  guides(color=F)+
  labs(x = "Experience (years)",
       y = "Wage",
       title = "Parallel Slopes Model",
       subtitle = "wage = 6.63+0.03 Experience-2.48 Female")+
  theme_classic(base_family = "Fira Sans Condensed", base_size=20)
```
]

### 2019-11-19 content/r/19-r-practice.Rmd
Fix typos on practice 19, HW 4 due Thursday not Tuesday
Update paper deadlines and data sources

- [<i class="fas fa-table"></i> `speeding_tickets.csv`](/data/speeding-tickets.csv)

@@

where
- `df` is the name of your dataframe
- `my_var` and `my_var2` are example variables^[As a bonus, you can try doing this with just one conditional command: `mutate_if(is.integer, factor)`. See our [Data Wrangling slides](https://metricsf19.classes.ryansafner.com/slides/05-slides#95) for refreshers of all the fancy `mutate()` possibilities!]

@@

where: 
  - `y` is the variable we are testing
  - `d` is the dummy variable

@@

Make a new variable called `Male` and save it in your dataframe.^[Hint: `mutate()` and define `Male` equal to the `ifelse()` function, which has three arguments: 1. the condition to test, 2. what to define "Male" as when the condition is `TRUE`, and 3. what to define "Male" as when the condition is `FALSE`.]

>>>>

- [<i class="fas fa-table"></i> `speeding_tickets.csv`](/data/speeding_tickets.csv)

@@

where

- `df` is the name of your dataframe
- `my_var` and `my_var2` are example variables^[As a bonus, you can try doing this with just one conditional command: `mutate_at(c("Black", "Hispanic", "Female", "OutTown", "OutState"),factor)`. See our [Data Wrangling slides](https://metricsf19.classes.ryansafner.com/slides/05-slides#95) for refreshers of all the fancy `mutate()` possibilities!]

@@

where:

  - `y` is the variable we are testing
  - `d` is the dummy variable

@@

Make a new variable called `Male` and save it in your dataframe.^[Hint: `mutate()` and define `Male` equal to `factor(ifelse())`. This makes the variable a `factor` (so we don't have to later), and the `ifelse()` function has three arguments: 1. `test =` the condition to test, 2. `yes =` what to define "Male" as when the condition is `TRUE`, and 3. `no = ` what to define "Male" as when the condition is `FALSE`.]

### 2019-11-19 static/r/19-r-practice-rmd.Rmd
Fix typos on practice 19, HW 4 due Thursday not Tuesday
Update paper deadlines and data sources

where
- `df` is the name of your dataframe
- `my_var` and `my_var2` are example variables^[As a bonus, you can try doing this with just one conditional command: `mutate_if(is.integer, factor)`. See our [Data Wrangling slides](https://metricsf19.classes.ryansafner.com/slides/05-slides#95) for refreshers of all the fancy `mutate()` possibilities!]

@@

### Part A
Make a new variable called `Male` and save it in your dataframe.^[Hint: `mutate()` and define `Male` equal to the `ifelse()` function, which has three arguments: 1. the condition to test, 2. what to define "Male" as when the condition is `TRUE`, and 3. what to define "Male" as when the condition is `FALSE`.]

>>>>

where
- `df` is the name of your dataframe
- `my_var` and `my_var2` are example variables^[As a bonus, you can try doing this with just one conditional command: `mutate_at(c("Black", "Hispanic", "Female", "OutTown", "OutState"),factor)`. See our [Data Wrangling slides](https://metricsf19.classes.ryansafner.com/slides/05-slides#95) for refreshers of all the fancy `mutate()` possibilities!]

@@

### Part A
Make a new variable called `Male` and save it in your dataframe.^[Hint: `mutate()` and define `Male` equal to `factor(ifelse())`. This makes the variable a `factor` (so we don't have to later), and the `ifelse()` function has three arguments: 1. `test =` the condition to test, 2. `yes =` what to define "Male" as when the condition is `TRUE`, and 3. `no = ` what to define "Male" as when the condition is `FALSE`.]
